Lab 2 Final Report: COVID-19 Case Rate vs Population Demographics and Mask Policy

### 2020-12-08 Jackson_Liu_Temlock_Zhang_Lab2_Final_Report.Rmd
Updated RMD per meeting discussion and added new file for Final Report

```{r variables, warning=FALSE, results='hide'}
df <- read.csv("covid-19.csv", header = TRUE)
df<-df%>%
  rename(case_rate_100k = 'Case.Rate.per.100000',
         population_density = 'Population.density.per.square.miles',
         mask_public='Mandate.face.mask.use.by.all.individuals.in.public.spaces',
         poverty_pct = "Percent.living.under.the.federal.poverty.line..2018.",
         unemployed_pct = "Percent.Unemployed..2018.",
         senior_pct = "X65."
         ) %>%
  select(State, case_rate_100k, population_density, mask_public, poverty_pct, 
         unemployed_pct, senior_pct)

head(df)
```

@@

**Research Question:** How is the COVID-19 case rate related to the distribution of population demographics within a state, and how does the effect of these demographics compare with the effect of mask-related policy decisions made by that state?

For this report, the investigation will be centered around the effect of population demographics and policy on the COVID-19 case rates across the United States (US), which is grouped by the 50 states plus the District of Columbia (D.C.). Within the report, this collection will be referred to as the "states", and each member a "state", inclusive of D.C..

@@

The first and primary phase of investigation involves how the COVID-19 case rate is related to key population demographics. For this, the key variables are COVID-19 case rate per 100,000 residents and the percentage population distribution of seniors (defined as 65 years old or older in the original dataset) within a state. The case rate per 100,000 was selected as the key output/dependent variable as it provides a standardized measure of the spread of COVID-19 across the states, and already takes into account the absolute population of the states. The senior percentage was selected as the key input/independent variable given that guidance had been released by the Centers for Disease Control and Prevention (CDC) that seniors belong to the age category for those at higher risk of COVID-19, and thus are a well studied and documented group. This is likely due to both the fact that seniors in this age group are more likely or had greater access to be tested given their categorization, and the fact that they would have a greater likelihood of exhibiting detectable COVID-19 related symptoms that would prompt them to get tested. An additional rationale for measuring the senior percentage is that they may also be more likely to contract COVID-19 at lower viral loads, making them more susceptible to the virus. Conversely, this may result in seniors being more wary of the threat of COVID-19, and taking additional precautions to prevent infection relative to other age categories, such as limiting social interactions and taking more preventative measures in terms of hygiene. These key variables will provide an initial understanding of the relationship between the spread of the virus and population demographics.

@@

Additionally, the final phase of investigation continues to expand on the examination of state-level population demographics via the population density variable (defined as the population density in square miles, population/square miles in the original dataset). A key factor to the spread of COVID-19 is the idea of social distancing, where the virus is considered to be more likely to spread when people are close in proximity. Thus, population density is measured as a conduit to indicate the level of proximity within each state and will operationalize the concept of social distancing. It should be noted however that the population density data may be limited in its ability to serve as a direct measurement of social distancing given the it lacks other contextualize information, such as the urban to rural ratio as later discussed in section 5.

@@

Secondly, note that there are many internal and external aspects of the selected variables that have not been included within the models. Just addressing the internal information that is lost, it can be seen that some of the information is not captured given the methods of operationalization discussed previously. Among others, the loss of information on the date of implementation within the mask policy variable strips out any contextual knowledge regarding the length of policy implementation, and precludes the identification of how long a specific policy was implemented. Assuming that this has some effect on case rate, this difference in length of time may play a part in the effect. There are also other external aspects that cannot be included, such as the rural to urban ratio as opposed to population density.

Finally, with regard to the policy variables, this report focuses on a form of mask mandate. Given this, it does not capture the effect of other policies that may or may not have a greater effect on the case rate (e.g., implementation of stay-at-home orders, closure of non-essential businesses, etc.). In justification, these choices were largely made due to there either being an appropriate amount of samples in each category (e.g., most states have implemented basic policies such as the closure of non-essential businesses and the implementation of stay-at-home orders, and thus these were not included given the lack of samples for the states that have not implemented them), or that data on other policies were simply not readily available in the dataset. The same argument can be applied in the selection of population demographics.

@@

In order to allow for more simplistic and direct interpretations when analyzing the effect of the input variables on the output, it was determined that all percentage variables (senior, poverty, and unemployment) would be transformed into rates per 100,000 to align with the case rate variable. It will also serve to ease the burden on the readers of this report as like-for-like scaled relationships can be drawn between the variables. Given that this transformation is constant for all values (scalar multiplication of 100,000 for the seniors variable given it is a ratio, and 1,000 for poverty and unemployment given they are percentages on a scale of 0-100%), it will have no effect on the distributions of the variables.

@@

For the first model, the relationship between the COVID-19 case rate per 100,000 and the population percentage distribution of seniors was analyzed.

@@

For model 3, policy related variables will be introduced for the analytics. The policy variable "Mandate face mask use by all individuals in public spaces" (renamed as mask_public) is selected as control variable to predict the case rate. Given it has been proven in external research that masks are effective in preventing the transmission of COVID-19, it is expected that the mask policy mandate in public would have an effect in reducing the case rate.

@@

```{r, message=FALSE, warning=FALSE}
df %>% 
  ggplot(aes(x = factor(mask_public_bool), y = case_rate_100k)) + 
  geom_point(aes(size=5, alpha=0.2), show.legend = FALSE) + 
  labs(
    title = 'Case Rate vs Mask Public Mandate', 
    x = 'Mask Mandate in Public', 
    y = 'Case Rate per 100,000')  +
  scale_x_discrete(labels=c("0","1"))

df %>%
  ggplot(mapping = aes(factor(mask_public_bool), case_rate_100k))+
  geom_jitter(shape=21, size=6, aes( fill = factor(mask_public_bool), alpha=0.5), width=0.1)+
  theme(legend.position = "none") +
  labs(
    title = 'Case Rate vs Mask Public Mandate', 
    x = 'Mask Mandate in Public', 
    y = 'Case Rate'
  ) +
  scale_x_discrete(labels=c("No","Yes"))

df %>%
  ggplot(aes(case_rate_100k))+
  geom_histogram(aes(y = ..density.., fill = factor(mask_public_bool)), position = 'identity', alpha = 0.5, color = 'white')+
  stat_density(geom = 'line', size=1, position = 'identity', aes(color = factor(mask_public_bool)))+
  scale_fill_brewer(palette = 'Dark2') +
  scale_color_brewer(palette = 'Dark2') +
  labs(
    title = 'Distribution of COVID-19 Case Rate',
    x = 'Covid-19 case rate per 100,000'
  ) +
  scale_fill_discrete(name='Mask public policy',labels=c('No', 'Yes'))+
  scale_color_discrete(name='Mask public policy',labels=c('No', 'Yes'))
```

@@

 - Phase 1: To investigate the direct relationship between COVID-19 case rate and the senior rate within a state
 - Phase 2: To build upon Phase 1 by including two additional variables of population demographics, poverty rate and unemployment rate, and investigating the overall effect on COVID-19 case rate
 - Phase 3: To build upon Phase 2 by adding a variable to measure the effect of the implementation of a mask-policy on COVID-19 case rate. A variable for population density was also added as an additional measure of population demographics

@@

Starting with the variable with the largest coefficient magnitude, it can be seen that with the implementation of a mask policy in a state there are 945 fewer cases per 100,000 compared to a state without any implementation of such policy. Given that research from leading health organizations shows that the wearing of masks helps to decrease the chance of the transmission of COVID-19, it could be suggested that a mask policy is effective in increasing the adoption of wearing masks among the state population. Next, it can be seen that there is a decrease of roughly 1 case per 100,000 with every 3 additional people unemployed per 100,000. This suggests that as unemployment increases, the case rate decreases, which could be a result of the fact that those who are unemployed may be less mobile as they are not commuting into work, lowering the likelihood of contracting the virus. Following unemployment, when examining the poverty variable there is actually an increase of roughly 1 case per 100,000 for every 5 additional people per 100,000 living under the federal poverty line. Although there is some intuition behind the unemployment and poverty rates having similar effects on case rate, it can be seen from the test of collinearity in subsection 2-2-3 that these two variables are not explaining the same effects. This difference is likely due to the fact that those living below the poverty line have less access to suitable living conditions and sanitary goods that help to alleviate the possibility of COVID-19 infection. As a caveat, it should be noted that these the poverty and unemployment figures are from a 2018 survey, so it does not reflect the full picture of the state demographics at the time of when the COVID-19 related data was captured. Finally, the effect of the senior rate is negative, with an additional 4 seniors per 100,000 translating to roughly 1 fewer case per 100,000. As touched upon in the introduction, this could be largely down to the fact that it is widely broadcasted that seniors fall within the high risk category in terms of having severe complications from COVID-19, and therefore they are more diligent in the precautions they take against contracting the virus.

>>>>

```{r variables, warning=FALSE, results='hide'}
df <- read.csv("covid-19.csv", header = TRUE)
df<-df%>%
  rename(case_rate_100k = 'Case.Rate.per.100000',
         population_density = 'Population.density.per.square.miles',
         mask_public='Mandate.face.mask.use.by.all.individuals.in.public.spaces',
         poverty_pct = "Percent.living.under.the.federal.poverty.line..2018.",
         unemployed_pct = "Percent.Unemployed..2018.",
         senior_pct = "X65."
         ) %>%
  select(State, case_rate_100k, population_density, mask_public, poverty_pct, 
         unemployed_pct, senior_pct)
```

@@

**Research Question:** How is the COVID-19 case rate related to the distribution of population demographics and policy within a state?

For this report, the investigation will be centered around the relationship of population demographics and policy on the COVID-19 case rates across the United States (US), which is grouped by the 50 states plus the District of Columbia (D.C.). Within the report, this collection will be referred to as the "states", and each member a "state", inclusive of D.C..

@@

The first and primary phase of investigation involves how the COVID-19 case rate is related to key population demographics. For this, the key variables are COVID-19 case rate per 100,000 residents and the senior rate (defined as 65 years old or older in the original dataset) within a state. The case rate per 100,000 was selected as the key output/dependent variable as it provides a standardized measure of the spread of COVID-19 across the states, and already takes into account the absolute population of the states. The senior percentage was selected as the key input/independent variable given that guidance had been released by the Centers for Disease Control and Prevention (CDC) that seniors belong to the age category for those at higher risk of COVID-19, and thus are a well studied and documented group. This is likely due to both the fact that seniors in this age group are more likely or had greater access to be tested given their categorization, and the fact that they would have a greater likelihood of exhibiting detectable COVID-19 related symptoms that would prompt them to get tested. An additional rationale for measuring the senior percentage is that they may also be more likely to contract COVID-19 at lower viral loads, making them more susceptible to the virus. Conversely, this may result in seniors being more wary of the threat of COVID-19, and taking additional precautions to prevent infection relative to other age categories, such as limiting social interactions and taking more preventative measures in terms of hygiene. These key variables will provide an initial understanding of the relationship between the spread of the virus and population demographics.

@@

Additionally, the final phase of investigation continues to expand on the examination of state-level population demographics via the population density variable (defined as the population density in square miles, population/square miles in the original dataset). A key factor to the spread of COVID-19 is the idea of social distancing, where the virus is considered to be more likely to spread when people are close in proximity. Thus, population density is measured as a conduit to indicate the level of proximity within each state and will operationalize the concept of social distancing. It should be noted however that the population density data may be limited in its ability to serve as a direct measurement of social distancing given it lacks other contextualize information, such as the urban to rural ratio as later discussed in section 5.

@@

Secondly, note that there are many internal and external aspects of the selected variables that have not been included within the models. Just addressing the internal information that is lost, it can be seen that some of the information is not captured given the methods of operationalization discussed previously. Among others, the loss of information on the date of implementation within the mask policy variable strips out any contextual knowledge regarding the length of policy implementation, and precludes the identification of how long a specific policy was implemented. Assuming that this has some relationship with case rate, this difference in length of time may play a part. There are also other external aspects that cannot be included, such as the rural to urban ratio as opposed to population density.

Finally, with regard to the policy variables, this report focuses on a form of mask mandate. Given this, it does not capture other policies that may or may not have other relationships with the case rate (e.g., implementation of stay-at-home orders, closure of non-essential businesses, etc.). In justification, these choices were largely made due to there either being an appropriate amount of samples in each category (e.g., most states have implemented basic policies such as the closure of non-essential businesses and the implementation of stay-at-home orders, and thus these were not included given the lack of samples for the states that have not implemented them), or that data on other policies were simply not readily available in the dataset. The same argument can be applied in the selection of population demographics.

@@

In order to allow for more simplistic and direct interpretations when analyzing the relationships of the input variables on the output, it was determined that all percentage variables (senior, poverty, and unemployment) would be transformed into rates per 100,000 to align with the case rate variable. It will also serve to ease the burden on the readers of this report as like-for-like scaled relationships can be drawn between the variables. Given that this transformation is constant for all values (scalar multiplication of 100,000 for the seniors variable given it is a ratio, and 1,000 for poverty and unemployment given they are percentages on a scale of 0-100%), it will have no effect on the distributions of the variables.

@@

For the first model, the relationship between the COVID-19 case rate per 100,000 and senior rate was analyzed.

@@

For model 3, policy related variables will be introduced for the analytics. The policy variable "Mandate face mask use by all individuals in public spaces" (renamed as mask_public) is selected as an input variable in model 3. Given it has been proven in external research that masks are effective in preventing the transmission of COVID-19, it is expected that the mask policy mandate in public is related to a decrease in case rate.

@@

```{r, message=FALSE, warning=FALSE}
df %>%
  ggplot(mapping = aes(factor(mask_public_bool), case_rate_100k))+
  geom_jitter(shape=21, size=6, aes( fill = factor(mask_public_bool), alpha=0.5), width=0.1)+
  theme(legend.position = "none") +
  labs(
    title = 'Case Rate vs Mask Public Mandate', 
    x = 'Mask Mandate in Public', 
    y = 'Case Rate per 100,000'
  ) +
  scale_x_discrete(labels=c("0","1"))
```

@@

 - Phase 1: To investigate the direct relationship between COVID-19 case rate and the senior rate within a state
 - Phase 2: To build upon Phase 1 by including two additional variables of population demographics, poverty rate and unemployment rate, and investigating the overall relationship on COVID-19 case rate
 - Phase 3: To build upon Phase 2 by adding a variable to measure the relationship of the implementation of a mask-policy on COVID-19 case rate. A variable for population density was also added as an additional measure of population demographics

@@

Starting with the variable with the largest coefficient magnitude, it can be seen that with the implementation of a mask policy in a state there are 945 fewer cases per 100,000 compared to a state without any implementation of such policy. Given that research from leading health organizations shows that the wearing of masks helps to decrease the chance of the transmission of COVID-19, it could be suggested that a mask policy is effective in increasing the adoption of wearing masks among the state population. Next, it can be seen that there is a decrease of roughly 1 case per 100,000 with every 3 additional people unemployed per 100,000. This suggests that as unemployment increases, the case rate decreases, which could be a result of the fact that those who are unemployed may be less mobile as they are not commuting into work, lowering the likelihood of contracting the virus. Following unemployment, when examining the poverty variable there is actually an increase of roughly 1 case per 100,000 for every 5 additional people per 100,000 living under the federal poverty line. Although there is some intuition behind the unemployment and poverty rates having similar relationships with case rate, it can be seen from the test of collinearity in subsection 2-2-3 that these two variables are not explaining the same effects. This difference is likely due to the fact that those living below the poverty line have less access to suitable living conditions and sanitary goods that help to alleviate the possibility of COVID-19 infection. As a caveat, it should be noted that these the poverty and unemployment figures are from a 2018 survey, so it does not reflect the full picture of the state demographics at the time of when the COVID-19 related data was captured. Finally, the coefficient of the senior rate is negative, with an additional 4 seniors per 100,000 translating to roughly 1 fewer case per 100,000. As touched upon in the introduction, this could be largely down to the fact that it is widely broadcasted that seniors fall within the high risk category in terms of having severe complications from COVID-19, and therefore they are more diligent in the precautions they take against contracting the virus.
